Species motif participation provides unique information about species risk of extinction

### 2022-06-10 manuscript/response_PRSB_round2.tex

        \us
        We are not convinced that collecting data on trophic level is easier than collecting entire food web data. If the Reviewer is referring to stable-isotope trophic levels, then we must point out that calculating trophic levels requires a good background understanding of the food sources for the focal species and rates of incorporation of different isotopes across trophic levels (and how these may differ between taxa). For a large community (e.g., 100 species), it could well be simpler to at least approximate the entire food web (giving trophic level, degree, and motifs simultaneously) than to calculate stable isotope-based trophic levels for each species. Indeed, the larger set of empirical food webs than comprehensive stable-isotope data suggests that food webs are easier to estimate at scale than trophic levels alone.

        To address this concern, we have added a brief statement in the introduction to make it clear to readers that researchers who have enough information to calculate in-degree and trophic level can also calculate motifs. Motif roles, therefore, provide additional information than degree and trophic level alone without requiring extra sampling.

        \them
        Third, I realize that this technique is already published on, however, some of the limitations of the technique may impact the utility of the results. I think the study would be greatly enhanced by a smaller comparison of a more traditional dynamic food web model approach and this approach – even just placed in the appendix. I note that the authors rely quite heavily on a paper on BioRxivs which has done something similar, so anticipate this wouldn’t be that difficult, but would be very convincing.

        \us We agree that a comparison between Bayesian and dynamical simulations could be interesting, however performing such a comparison is beyond the scope of this manuscript. As the Reviewer notes, a subset of the authors of this manuscript are working on an analysis of the relationships between motif participation and persistence in a dynamical framework (currently under review). We do not believe that including the material from this second paper in an appendix is justified when there are already a published comparisons of extinctions simulated by Bayesian and dynamical approaches (\citealp[]{Eklof2013a}). That paper focused on comparing the number and identity of secondary extinctions between the two approaches, and additionally evaluated which functional response of consumers to loss of prey that most accurately predicted the secondary extinctions from a fully-fledged dynamical model. We use this information in the current manuscript and make our simulations with the functional form that was shown to most accurately predict the same secondary extinctions as the dynamical model. We have now calrified this in the manuscript (line XX). However, we would also like to point out that there are no guarantees that advanced (as in parameter intensive) dynamical models describe the ecological reality better, as comparisons with extensive empirical extinction scenarios are still lacking.

        In addition, we believe that the Reviewer has over-estimated our reliance on the BioRxiv pre-print currently under review. Our previous draft had only three references to this pre-print: when describing the generation of simulated networks, when noting that degree and motif participation are not independent, and when briefly comparing the dynamical and Bayesian results in the discussion.
        We have now removed two of these references.
        We present the full network-generation methods in \emph{Appendix 1} and, to avoid confusion by referring to results presented in the preprint, have softened our language to say that degree and motif participation \emph{may} be correlated. 
        The latter is not a major point in our manuscript; we mention it only to emphasize that controlling for the total number of motifs in which a species appears does not necessarily control for degree. 
        Thus, we see some overlap in the associations of persistence with degree and with normalised motifs.

@@

        \them
        Line 87 – 89: similar to comment above – I think the idea of global and local food web properties has not been introduced at all, nor the classic food web properties (which ones are being measured? Only the four specific motifs thought to contribute to stability are mentioned).

        \us We have revised the introduction to more clearly introduce these concepts.

        \them
        Line 92 – 95: this seems to be a flaw to the Bayesian network simulations – we know that mutual predation happens in food webs – especially if the food web isn’t explicitly size structured, and when nested within the broader food web context these feedback loops may have stabilizing, or destabilizing effects.
        \smallskip
        Line 149 – 152: these seem like large limitations to me.

        \us In revising the introduction, we now introduce the Bayesian network approach and its limitations in the methods section. We want to stress that the Reviewer is clearly correct that these are important limitations and we have further clarified that in the text. At the same time we have clarified the advantages that the Bayesian network approach provides.
        First, we clearly state that the lack of top-down effects in the Bayesian networks is a limitation. However, we also point out that in prior studies Bayesian simulations have captured most secondary extinctions identified using dynamical simulations. This suggests that, in many cases, bottom-up effects are the main drivers of secondary extinction. We also emphasize the major gain in computational speed when using the Bayesian approach, which allows us to investigate different levels of disturbance rather than simple species removals that needs to be repeated numerous times. Finally, we have also added text to clarify how the process of making the food webs acyclic can affect the results, but that food web robustness rarely is affected.

@@

        \us We are not aware of any other motif-focused papers calling this motif ``exploitative competition''. Our terminology follows the nomenclature from the introduction of motifs to food webs~\citep{Stouffer2007}.
        Note that ``direct competition'' refers simply to two species which consume the same prey and is agnostic as to whether this is exploitative or interference competition. 
        As exploitative and interference competition are not the same ecologically, we have maintained the conventional motif terminology so as not to imply an overly specific interpretation.

        \them
        Line 334: in-degree and trophic level are not sufficient to explain the trends, yet there is a strong negative correlation between apparent competition and in-degree. This seems contradictory.

        \us The Reviewer is correct that there is a strong  correlation between apparent competition and in-degree, as we state in the text. It is also true that apparent competition and in-degree show opposite relationships with persistence, such that the apparent competition-persistence relationship could be described by the in-degree-persistence relationship. However, as detailed later in the paragraph, this ignores the strong correlation between apparent competition and trophic level which would suggest that persistence should show different relationships to apparent competition. This means that either trophic level is unrelated to persistence (which we know is not true) or some other factor is over-riding the ``effect" (there is no clear causality from trophic level to motifs or vice versa) of trophic level on apparent competition.

        We have revised this paragraph with particular focus on these relationships and hope that clarifies our reasoning . %to stop readers from jumping to conclusions before reading the entire section.

### 2022-06-13 manuscript/reorder_main_text.tex
Updates from Overleaf

\author{Anna \r{A}kesson$^{1*}$, Alyssa R. Cirtwill$^{2*\dagger}$, Kate L. Wootton$^{3}$, Anna Ekl\"{o}f$^{1}$} 
\date{\small$^1$Department of Theoretical Biology, Chemistry, and Physics\\ 
Link\"{o}ping University\\
Link\"{o}ping, Sweden\\
\medskip
\small$^2$Department of Agricultural Sciences\\
University of Helsinki\\
Helsinki, Finland\\
\medskip
\small$^3$ BioFrontiers Institute\\
University of Colorado, Boulder\\
U.S.A.
\medskip
* Joint first authors
\medskip
$^\dagger$ Corresponding author\\
}

@@

    %\subsection*{2 Global and local network properties are known to affect stability}
    How disturbances, such as population decline or extinction, lead to loss of other species via direct and indirect effects (i.e., secondary extinctions) has for long been vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    Several global (whole-network) as well as local (single-species) properties are known to influence how species respond to disturbances. Two well-known, influential global properties are network size, the number of species present in the network, and network connectance, the number of realized feeding interactions.
    Based on simulations of random networks, large networks were initially thought to be less stable~\citep{May1972}.
    The indisputable presence of large, stable, empirical networks has spurred research into how other structural characteristics, including connectance, may allow large numbers of species to persist~\citep{Dunne2002}.
    Highly-connected networks may be more resistant to secondary extinctions because consumers, on average, have access to a larger number of prey species and can switch from a declining or extinct prey to another resource~\citep{Dunne2002, Eklof2006,Baumgartner2015}. However, a well-connected network also includes more pathways through which a disturbance can spread~\citep{Dunne2004, Vieira2015}. Depending on the strength and arrangements of interactions, this means that species may be impacted through several paths, amplifying the original disturbance and causing more secondary extinctions.
    The net effect of connectance on extinction therefore likely depends on finer-scale network structure, as well as the type and severity of disturbance.

    Species properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). This can be referred to local properties of the network.
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, and secondary extinctions due to reduced plant populations are particularly detrimental \citep{Curtsdotter2011}. %including secondary extinction due to reduced plant populations.
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    This since, as well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may, for example, be at greater risk of extinction than a specialist which consumes a highly generalist resource that is more unlikely to go extinct.

    %\subsection*{3 Motif participation may similarly affect species' particular extinction risk}
    It is possible to capture information about how a consumer and its prey fit together within a community by defining participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. 
    Such motif profiles of species are easily calculated once we know the network structure and require no more information than is used to calculate degree or trophic level.
    These `motif participation' roles capture information about species' direct as well as indirect interactions, incorporating larger-scale network structure into species-level descriptions~\citep{Cirtwill2015a,Cirtwill2018FoodWebs}. We refer to these motifs as meso-scale structures, to emphasise their role as structures between global (network) and local (species) properties.  A species' motif participation profile shows the second step indirect interactions the species are affected by, and thereby capture important structures that neither global or local properties do.  
    Therefore, by allowing researchers to distinguish between, for example, consumers of specialists and generalists, motif participation may complement degree and trophic level to more fully predict species' extinction risk.

@@

% Severity of disturbance
    As mentioned above, the severity of a disturbance may interact with network structure to affect secondary extinctions.
    A mild disturbance (e.g., leading to a small increase in the probability of basal resources going extinct) is intuitively likely to cause fewer secondary extinctions than a severe disturbance, and the difference in effects can depend on network structure~\citep{Baumgartner2015}.
    Empirical studies have shown that disturbances to plant species can affect some consumers more strongly than others  \citep{byrnes2011climate}, that the strength of disturbance affects which consumer species are most strongly affected~\citep{detmer2021variation,carnell2020more}, and that the resilience and return time of the ecological community can be strongly influenced by disturbance severity \citep{rydgren2004disturbance}. The severity disturbance therefore clearly has an effect on how the ecological community will respond.

%\subsection*{Here we test...}
    Here, we are interested in whether a consumer's extinction risk of going secondarily extinct depends on 1) the motifs in which the species participates, 2) the motif profile of the full food web and 3) the severity of the disturbance to the basal level. %We address this in scenarios with various levels of disturbance targeting the basal level.
    In addition, we want to understand how the information provided by motifs is related to other global and local food web properties in order to better synthesize information provided by different scales of network structure.
    We use synthetic networks and model secondary extinctions using Bayesian networks \citep{Eklof2013, Haussler2020}. As Bayesian networks are computationally efficient (see Methods), we are able to simulate a wide range of disturbance intensities and test whether meso-scale network structure also interacts with severity of the disturbance.
    We focus our analyse on the four network motifs that are over-represented in empirical networks and additionally have been shown to contribute to network stability; the three-species chain, omnivory, apparent competition, and direct competition \citep{Stouffer2007, Borrelli2015a, giling2019plant}.

% % OLD INTRODUCTION BELOW
% \section*{Introduction} %Old, for comparison

%     %\subsection*{1 Why are bottom-up effects important for understanding stability?}
%      The plant community is the foundation upon which the myriad species in terrestrial food webs depend for their survival. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}.
%      The structure of the plant community affects stability across all levels of ecosystem organization, which is crucial for sustaining a healthy energy flow in the ecosystem as a whole~\citep{proulx2010diversity, scherber2010bottom, Rosenblatt2016}. 
%      For example, the loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}.

%      %\subsection*{2 Bayesian networks are a fantastic way to model bottom-up effects}
%     How population decline or extinction of one species leads to loss of other species via direct and indirect effects (i.e., secondary extinctions) is a vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}. Traditionally, there are two main approaches for studying secondary extinctions. First, there are topological models, based only on food web structure \citep{dunne2009cascading}. Here, extinctions only affect other species in the network once a consumer has lost all of its prey and therefore must go extinct. The second approach uses dynamical models which explicitly simulate population dynamics using a system of differential equations \citep{binzer2011susceptibility}. Dynamical models take changes in prey or predator densities into account when calculating the densities of other species in the network. 
%     This additional detail means that realistic processes such as indirect interactions can be included, but also means that dynamical models are more parameter intensive, and simulations are more time-consuming, than topological models.

%     A middle‐ground approach for simulating secondary extinctions is to use Bayesian networks \citep{Eklof2013}. 
%     In this framework, a consumer's probability of extinction depends on the fraction of resources lost and on a baseline probability of extinction which captures causes unrelated to the network structure (e.g., disease, stochastic extinction of small populations).
%     Modelling Bayesian networks is fast and computationally efficient, allowing analysis of networks with high species richness \citep{Haussler2020}. 
%     As consumers do not affect the extinction probabilities of their resources, Bayesian networks can only capture bottom-up effects of disturbances within networks. 
%     However, Bayesian networks do capture the majority of secondary extinctions captured by dynamical models \citep{Eklof2013}.
%     This emphasizes the importance of bottom-up effects in food webs and makes Bayesian networks the ideal framework to use when studying them.

%     %\subsection*{3 Global scale network properties are known to affect stability}
%     Using Bayesian networks, researchers can explore how the structures of realistically-large networks shape their responses to disturbance.
%     The global (whole-network) structural properties of ecological networks have implications for their functioning \citep{Petchey2002}, stability \citep{Allesina2012} and robustness to disturbances \citep{Dunne2002, Eklof2006}. 
%     For example, more highly-connected networks are more resistant to secondary extinctions \citep{Dunne2002, Eklof2006} and a modular organization, where some groups of species are tightly interconnected with limited connections between groups, can limit the propagation of disturbances \citep{Krause2003, Teng2004, Gilarranz2017}.

%     %\subsection*{4 meso-scale nw propoerties also affect stability}
%     Structural properties on the local (single-species) scale are also known to affect response to disturbance. 
%     Species with high trophic levels (long paths to basal resources), low in-degree (few prey), and low abundances are generally at higher risks of secondary extinction \citep{binzer2011susceptibility}. 
%     However, in-degree and trophic level are both relatively coarse descriptions of a species' role within its food-web~\citep{Cirtwill2018FoodWebs}. 
%     For example, a specialist may not necessarily have a high risk of extinction if it consumes a basal resource which is unlikely to be affected by disturbance. 
%     On the other hand, a consumer with low trophic level may be at high risk of extinction if it consumes prey which are themselves very likely to be affected by disturbance.

%     %\subsection*{5 Motif participation may similarly affect species' particular extinction risk}
%     One way to incorporate more nuance into descriptions of how species fit into their communities is to define species' roles based on their participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. This can be referred to as meso-scale properties and describe how a species is interacting with its closest neighbours. As such, meso-scale properties is thus one a scale between local properties and global properties. Importantly, these `motif participation' roles capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions~\citep{Cirtwill2015a}. 
%     This means that two species with the same in-degrees and trophic levels can still participate in different sets of motifs if, for example, one interacts with specialists and the other interacts with generalists~\citep{Cirtwill2018FoodWebs}. 
%     This extra detail may be useful when predicting  extinction risk since, at the network level, the frequencies of different motifs are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}. 
%     We therefore expect that both the motif structure of a network and species' participation in that structure should be related to risks due to disturbance to basal resources.

%           %\subsection*{Here we test...}
%     Specifically, we are interested in whether a consumer's extinction risk depends on 1) the motifs in which a focal species participates and 2) the motif profile of the full food web. 
%     In addition, we want to understand how the information provided by motifs is related to classical global and local food web properties in order to better synthesize information provided by different scales of network structure.
%     Of the 13 three-species motifs that can occur in empirical food webs, there are four motifs that are over-represented in empirical networks and are thought to contribute to stability \citep{Stouffer2007, Borrelli2015a, giling2019plant}. These four motifs, the three-species chain, omnivory, apparent competition, and direct competition, are also the only ones which can occur in the acyclic networks required for Bayesian network simulations~\citep{Eklof2013}. 
%     Therefore, we focus on these four particularly important motifs.

@@

        Traditionally, there are two main approaches for studying secondary extinctions. First, there are topological models, based only on food web structure \citep{dunne2009cascading}. Here, extinctions only affect other species in the network once a consumer has lost all of its prey and therefore must go extinct. 
        The second approach uses dynamical models which explicitly simulate population dynamics using a system of differential equations \citep{binzer2011susceptibility}. Dynamical models take changes in prey or predator densities into account when calculating the densities of other species in the network. 
        This additional detail means that realistic processes such as indirect interactions can be included, but also means that dynamical models are extensively more parameter intensive, and simulations are much more time-consuming, than topological models.

        A middle‐ground approach for simulating secondary extinctions is to use Bayesian networks \citep{Eklof2013, Haussler2020}. 
        In this framework, a consumer's probability of extinction depends on the fraction of resources lost and on a baseline probability of extinction which captures causes affecting the respective species extinction probability that are unrelated to the network structure (e.g., disease, stochastic extinction of small populations).
        Modelling Bayesian networks is fast and computationally efficient, allowing analysis of networks with high species richness \citep{Haussler2020}. 
        This increased speed allows us to simulate a range of disturbance severity rather than single species removals as in many previous studies~\citealp[e.g.,]{Memmott2004,Staniczenko2010,Dunne2004}.
        Moreover, Bayesian network simulations are less parameter-intensive than dynamical models, meaning that they are less sensitive to assumptions made in the modelling process~\citep{Eklof2013}.

       % In Bayesian networks consumers do not affect the extinction probabilities of their resources, and as such we can here only capture bottom-up effects of disturbances.  However, Bayesian networks do capture the majority of secondary extinctions captured by dynamical models \citep{Eklof2013}. Moreover, bottom-up effects are often strong in empirical systems. For example, bottom-up effects regulate zoo-plankton communities in freshwater lakes, regulating zoo-plankton communities \citep{li2020bottom} and control abundances and biomass of vertebrate species in several terrestrial systems \citep{Dobson2009food, Mduma1999food, Georgiadis2007}.

	%\subsection*{Calculating persistence}	
    % basic idea behind BN modeling
        Bayesian networks can be used to represent species and their feeding interactions as the nodes and links of a network, respectively (see Appendix \emph{S2} for full details).
        In our setting, each species (node) is a Bernoulli random variable, being either extant or extinct. 
        The probability that a particular species is extinct is a function of the states of its resources.

		% probability of extinction equation
		%Consider a species with $n$ resource species. 
		Before taking food-web structure into account, all species already have a baseline probability of extinction, $\pi$. 
		$\pi$ represents the risk of going extinct due to diseases, stochastic extinction of small populations, or other factors not related to the food web.
		A species probability of extinction also depends on the state of its resources. 
        If the focal species is a consumer with $n$ resources, of which $k$ have gone extinct, its probability of extinction depends on $\pi$ and on the fraction $f = k/n$.

@@

		If all resource species of a consumer \textit{C} are present ($f = 0$), the extinction probability will be $P(\lnot C|f) = \pi$. 
		Similarly, if all resources are extinct ($f = 1$), the extinction probability $P(\lnot C|f)$ will be equal to 1 and the consumer will go extinct.
		If the fraction $f$ is not 1 nor 0, the non-linear sigmoid curve will determine the response of the consumer, which will be more severe if the fraction resources lost is high and less severe if the fraction is small. This will in turn affect the extinction probability $P(\lnot C|f)$.
		For basal resources, we assume that abiotic resources are always available and $f=0$. Following Equation \ref{betafunc}, the extinction will only depend on $\pi$.
		Note that we here assume that only the fraction, and not the identity, of the resources lost is important. 
		However, the model can easily be modified to incorporate weighted resource interactions \citep[see][]{Eklof2013}.

		% how we calculate persistences 
        Beginning with basal resources and working systematically up through the food web, we use Equation \ref{betafunc} to calculate each species' $i$ probability of extinction $P(\lnot i|f)$.
        Using this probability, we then perform Bernoulli trials to determine whether the species goes extinct or not. 
        Species are considered extinct if a random number drawn from a uniform distribution $[0-1]$ is less than $P(\lnot i|f)$.

        These simulated extinctions are used when calculating the fraction of resources lost $f$ for each consumer.
        This allows us to update Equation~\ref{betafunc} with information on extinctions of a focal consumer's resources and simulate bottom-up effects. As this method is probabilistic, we repeat these calculations 100 000 times per network, with unique random draws.
	    For each species, we then define probability of persistence as the fraction of simulations in which the species persists. There are methods for solving Bayesian networks exactly \citep{Eklof2013}.
	    However, for larger networks the described numerical evaluation is highly efficient and produces close to identical results \citep{Haussler2020}.

		% BN limitations
		Modelling species extinctions in food webs using Bayesian networks has two main limitations. First, in Bayesian networks consumers do not affect the extinction probabilities of their resources, and as such we can here only capture bottom-up effects of disturbances. However, Bayesian networks has been shown to capture the majority of secondary extinctions produced fully-fledged dynamical models \citep{Eklof2013}. \cite{Eklof2013} analyzed which functional form of consumers response to loss of resources that most accurately predicted the secondary extinctions realized in the dynamical model, and we here use that functional form in our analyzes.  Moreover, bottom-up effects are often the main force of secondary extinctions in empirical systems. For example, bottom-up effects regulate zoo-plankton communities in freshwater lakes, regulating zoo-plankton communities \citep{li2020bottom} and control abundances and biomass of vertebrate species in several terrestrial systems \citep{Dobson2009food, Mduma1999food, Georgiadis2007}. Second, due to the strict bottom-up approach of Bayesian networks the networks must be acyclic. Fortunately, \cite{Allesina2009functional} proved that cyclic and acyclic versions of a network have the same properties in terms of secondary extinctions and the effect on network robustness are very limited. Thus, making the networks acyclic in order to use the Bayesian network approach will not strongly affect extinction scenarios.

    \subsection*{Disturbance scenarios}
        In our baseline scenario, all species had an extinction probability of $\pi = 0.1$. 
		To simulate various threats to basal resources, making some species more vulnerable to extinction, we increase the affected basal species' probability of extinction from $0.10$ to $0.5$, in steps of $0.08$. 
		The highest disturbance level, $\pi = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi=0.1$ in all cases.

@@

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw}. 
		To capture a range of plausible network architectures, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details and a comparison to empirical networks, see \emph{Appendix S1}.
		We removed any biologically unlikely networks (Appendix \emph{S1}) and replaced them with new simulated networks, repeating the process until we obtained 100 suitable networks in each combination of size and connectance.

    	To make the simulated networks compatible with the Bayesian simulation framework, all networks were rendered acyclic following~\citet{Allesina2009functional} and species topologically sorted from lowest to highest trophic level (Appendix \emph{S1}). These processing steps allow the strictly bottom-up calculation of probability of persistence.

>>>>

\author{Alyssa R. Cirtwill$^{1*\dagger}$, Anna \r{A}kesson$^{2*}$, Kate L. Wootton$^{3}$, Anna Ekl\"{o}f$^{2}$} 
\date{
\small$^1$Spatial Foodweb Ecology Group\\
Research Centre for Ecological Change\\
Organismal and Evolutionary Biology Research Programm\\
Faculty of Biological and Environmental Sciences\\
University of Helsinki
\medskip
\small$^2$Department of Theoretical Biology, Chemistry, and Physics\\ 
Link\"{o}ping University\\
Link\"{o}ping, Sweden\\
\medskip
\small$^3$ BioFrontiers Institute\\
University of Colorado, Boulder\\
U.S.A.
\medskip
* Joint first authors
\medskip
$^\dagger$ Corresponding author\\
}

@@

    %\subsection*{2 Global and local network properties are known to affect stability}
    How disturbances, such as population decline or extinction, lead to loss of other species via direct and indirect effects (i.e., secondary extinctions) has for long been vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    Several global (whole-network) as well as local (single-species) properties are known to influence how species respond to disturbances. Two well-known, influential global properties are network size, the number of species present in the network, and network connectance, the number of realized feeding interactions.
    Based on simulations of random networks, large networks were initially thought to be less stable~\citep{May1972}.
    The indisputable presence of large, stable, empirical networks has spurred research into how other structural characteristics, including connectance, may allow large numbers of species to persist~\citep{Dunne2002}.
    Highly-connected networks may be more resistant to secondary extinctions because consumers, on average, have access to a larger number of prey species and can switch from a declining or extinct prey to another resource~\citep{Dunne2002, Eklof2006,Baumgartner2015}. However, a well-connected network also includes more pathways through which a disturbance can spread~\citep{Dunne2004, Vieira2015}. Depending on the strength and arrangements of interactions, this means that species may be impacted through several paths, amplifying the original disturbance and causing more secondary extinctions.
    The net effect of connectance on extinction therefore likely depends on finer-scale (species-level) network structure, as well as the type and severity of disturbance.

    Species-level properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    These can also be referred to as local properties of the network.
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, and disturbances to plant populations cause particularly many econdary extinctions \citep{Curtsdotter2011}. %including secondary extinction due to reduced plant populations.
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    A more detailed perspective is desirable since, as well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may be at greater risk of extinction than a specialist which consumes a highly generalist resource that is more unlikely to go extinct (Fig.~\ref{fig:concept}).

    %\subsection*{3 Motif participation may similarly affect species' particular extinction risk}
    It is possible to capture information about how a consumer and its prey fit together within a community by defining participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. 
    Such motif profiles of species are easily calculated once we know the network structure and require no more information than is used to calculate degree or trophic level.
    These `motif participation' roles capture information about species' direct as well as indirect interactions, incorporating larger-scale network structure into species-level descriptions~\citep{Cirtwill2015a,Cirtwill2018FoodWebs}.
    We refer to these motifs as meso-scale structures to emphasise their role as structures between global (network-level) and local (species-level) properties.  
    A species' motif participation profile shows the second-step indirect interactions the species are affected by, thereby capturing important structures that are missed by global and local properties.  
    Therefore, by allowing researchers to distinguish between, for example, consumers of specialists and generalists, motif participation may complement degree and trophic level to more fully predict species' extinction risk.

@@

% Severity of disturbance
    As mentioned above, the severity of a disturbance may interact with network structure to affect secondary extinctions.
    A mild disturbance (e.g., leading to a small increase in the probability of basal resources going extinct) is intuitively likely to cause fewer secondary extinctions than a severe disturbance, and the difference in effects can depend on network structure~\citep{Baumgartner2015}.
    Empirical studies have shown that disturbances to plant species can affect some consumers more strongly than others \citep{byrnes2011climate}, that the strength of disturbance affects which consumer species are most strongly affected~\citep{detmer2021variation,carnell2020more}, and that the resilience and return time of the ecological community can be strongly influenced by disturbance severity \citep{rydgren2004disturbance}. 
    The severity of disturbance therefore clearly has an effect on how the ecological community will respond.

%\subsection*{Here we test...}
    Here, we are interested in whether a consumer's extinction risk of going secondarily extinct depends on 1) the motifs in which the species participates, 2) the motif profile of the full food web and 3) the severity of the disturbance to the basal level. %We address this in scenarios with various levels of disturbance targeting the basal level.
    In addition, we want to understand how the information provided by motifs is related to other global and local food web properties in order to better synthesize information provided by different scales of network structure.
    We use synthetic networks and model secondary extinctions using Bayesian networks \citep{Eklof2013, Haussler2020}. 
    As Bayesian networks are computationally efficient (see Methods), we are able to simulate a wide range of disturbance intensities and test whether meso-scale network structure also interacts with severity of the disturbance.
    We focus our analyse on the four network motifs that are over-represented in empirical networks and additionally have been shown to contribute to network stability: the three-species chain, omnivory, apparent competition, and direct competition \citep{Stouffer2007, Borrelli2015a, giling2019plant}.

@@

        Traditionally, there are two main approaches for studying secondary extinctions. First, there are topological models, based only on food web structure \citep{dunne2009cascading}. Here, extinctions only affect other species in the network once a consumer has lost all of its prey and therefore must go extinct. 
        The second approach uses dynamical models which explicitly simulate population dynamics using a system of differential equations \citep{binzer2011susceptibility}. Dynamical models take changes in prey or predator densities into account when calculating the densities of other species in the network. 
        This additional detail means that realistic processes such as indirect interactions can be included, but also means that dynamical models are highly parameter intensive and simulations are much more time-consuming than topological models.

        A middle‐ground approach for simulating secondary extinctions is to use Bayesian networks \citep{Eklof2013,Haussler2020}. 
        In this framework, a consumer's probability of extinction depends on the fraction of resources lost ($f = k/n$ for a species with $n$ resources, of which $k$ have gone extinct) and on a baseline probability of extinction ($\pi$) which captures factors affecting the focal species' extinction probability that are not related to the network structure (e.g., disease, stochastic extinction of small populations).
        Modelling Bayesian networks is fast and computationally efficient, allowing analysis of networks with high species richness \citep{Haussler2020}. 
        This increased speed allows us to simulate a range of disturbance severity rather than single species removals as in many previous studies (e.g.,~\citealp[]{Memmott2004,Staniczenko2010,Dunne2004}).
        Moreover, Bayesian network simulations are less parameter-intensive than dynamical models, meaning that they are less sensitive to assumptions made in the modelling process~\citep{Eklof2013}.

@@

		If all resource species of a consumer \textit{C} are present ($f = 0$), the extinction probability will be $P(\lnot C|f) = \pi$. 
		Similarly, if all resources are extinct ($f = 1$), the extinction probability $P(\lnot C|f)$ will be equal to 1 and the consumer will go extinct.
		If the fraction $f$ is not 1 nor 0, the non-linear sigmoid curve will determine the response of the consumer, which will be more severe if the fraction resources lost is high and less severe if the fraction is small. This will in turn affect the extinction probability $P(\lnot C|f)$.
		For basal resources, we assume that abiotic resources are always available and $f=0$. 
		Following Equation \ref{betafunc}, the extinction probability of a basal resource will only depend on $\pi$.
		Note that we here assume that only the fraction, and not the identity, of the resources lost is important. 
		However, the model can easily be modified to incorporate weighted resource interactions \citep[see][]{Eklof2013}.

		% how we calculate persistences 
        Beginning with basal resources and working systematically up through the food web, we use Equation \ref{betafunc} to calculate each species' $i$ probability of extinction $P(\lnot i|f)$.
        Using this probability, we then perform Bernoulli trials to determine whether the species goes extinct or not. 
        Species are considered extinct if a random number drawn from a uniform distribution $[0-1]$ is less than $P(\lnot i|f)$.
        These simulated extinctions are used when calculating the fraction of resources lost $f$ for each consumer.
        This allows us to update Equation~\ref{betafunc} with information on extinctions of a focal consumer's resources and simulate bottom-up effects. As this method is probabilistic, we repeat these calculations 100 000 times per network, with unique random draws.
	    For each species, we then define probability of persistence as the fraction of simulations in which the species persists. There are methods for solving Bayesian networks exactly \citep{Eklof2013}.
	    However, for larger networks the described numerical evaluation is highly efficient and produces close to identical results \citep{Haussler2020}.

		% BN limitations
		Modelling species extinctions in food webs using Bayesian networks has two main limitations. 
		First, in Bayesian networks consumers do not affect the extinction probabilities of their resources and as such we can here only capture bottom-up effects of disturbances. 
		However, Bayesian networks has been shown to capture the majority of secondary extinctions produced fully-fledged dynamical models \citep{Eklof2013}.
		\cite{Eklof2013} analyzed which functional form of consumers response to loss of resources that most accurately predicted the secondary extinctions realized in the dynamical model, and we here use that functional form in our analyzes.  
		Moreover, bottom-up effects are often the main force of secondary extinctions in empirical systems. 
		For example, bottom-up effects regulate zoo-plankton communities in freshwater lakes, regulating zoo-plankton communities \citep{li2020bottom} and control abundances and biomass of vertebrate species in several terrestrial systems \citep{Dobson2009food, Mduma1999food, Georgiadis2007}. 
		Second, due to the strict bottom-up approach of Bayesian networks the networks must be acyclic.
		Fortunately, \cite{Allesina2009functional} proved that cyclic and acyclic versions of a network have the same properties in terms of secondary extinctions and the effect on network robustness are very limited. 
		Thus, making the networks acyclic in order to use the Bayesian network approach will not strongly affect extinction scenarios.
		Moreover, the cyclic motifs which are excluded from an acyclic network are rare in empirical networks~\citep{Stouffer2007} and therefore unlikely to strongly affect extinction risk.

    \subsection*{Disturbance scenarios}

        In our baseline scenario, all species had an extinction probability of $\pi = 0.1$. 
		To simulate various threats to basal resources, making some species more vulnerable to extinction, we increase basal species' probability of extinction from $0.10$ to $0.5$, in steps of $0.08$. 
		The highest disturbance level, $\pi = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi=0.1$ in all cases.

@@

		We generated a realistic set of simulated networks based on the niche model~\citep{Williams2000,Stouffer2007} using the function ``nichemodel'' within the Julia~\citep{Bezanson2017julia} package \emph{BioEnergeticFoodWebs}~\citep{bioenergfw}. 
		To capture a range of plausible network architectures, we simulated networks with sizes (S) ranging from 50 to 100 species (in steps of 10) and connectances (C) ranging from 0.02 to 0.18 (in steps of 0.04). 
		For full details and a comparison to empirical networks, see \emph{Appendix S1}.
		We removed any biologically unlikely networks (Appendix \emph{S1}) and replaced them with new simulated networks, repeating the process until we obtained 100 suitable networks in each combination of size and connectance.
		To make the simulated networks compatible with the Bayesian simulation framework, all networks were rendered acyclic following~\citet{Allesina2009functional} and species topologically sorted from lowest to highest trophic level (Appendix \emph{S1}). 
		 These processing steps allow the strictly bottom-up calculation of probability of persistence.
